Social inequality in mortality among adults and elderly in northern Sweden 1851--2013\footnote{'Socoeconomic inequalities and mortality I: Europe'. 41st Annual Meeting of the Social Science History Association, Chicago, November 17--20, 2016.}

### 2016-11-19 slides.Rnw
fix nice presentation

\begin{frame}{Aims and questions}

@@

\begin{frame}{The Skellefteå and Umeå regions in Sweden}

<<umpop>>=
ume1900 <- with(obs[obs$region == "ume", ],
                sum(birthdate + enter < 1900 &
                    birthdate + exit > 1900))
ume1950 <- with(obs[obs$region == "ume", ],
                sum(birthdate + enter < 1950 &
                    birthdate + exit > 1950))
@

\includegraphics[height=8cm]{myfigs/skum17}\includegraphics[height=8cm]{myfigs/sweden}

\begin{frame}{The Skellefteå region}

\begin{itemize}  
\item Selection of
parishes surrounding the town of Skellefteå. 
\item Town founded in 1843 but with a
very small population during the first decades.
\item The majority of the
population in rural villages and hamlets, getting their
livelihood from agricultural production. 
\item 20th century: \emp{industrialisation}, population \emp{increase in the town},
\emp{diversified economy}. 
\end{itemize}

\begin{frame}{The Umeå region}

\begin{itemize}
\item \emp{Town} founded in \emp{1622}
\begin{itemize}
\item  Substantially \emp{larger} than Skellefteå
\item \emp{administrative} centre of the county Västerbotten
\item \emp{Military} regiments.
\item \emp{Schools}
\item \emp{University} from \emp{1963--65}.  
\end{itemize}  
\item The rural part similar to the one in the Skellefteå region.
\end{itemize}

@@

\begin{itemize}
\item \emp{Swedish church book} records (1851--1950).
  \begin{itemize}
\item digitized by the \emp{Demographic Data Base} at Umeå University, 
\begin{itemize}  
   \item \emp{\sc Popum} and 
   \item \emp{\sc Poplink}
\end{itemize}    
\item  \url{http://www.cedar.umu.se/}
\end{itemize}
\item The \emp{Linnaeus} data base (1960--2013)
  \begin{itemize}
\item Census data 1960, 1965, \ldots, 1990
\item Population registers ({\sc Lisa}, 1986--2013)
  \begin{itemize}
    \item \emp{Income}
    \item \emp{Education}  
  \end{itemize}    
 \item Death information from \emp{National Board of Health and Welfare}
   \begin{itemize}
\item Death dates from 1 January 1961 to 31 December 2013.
  \item \emp{Causes of death} ({\sc Ulorsak}, ``main cause of death'')
   \end{itemize}
\end{itemize}
\end{itemize}

@@

@@

\begin{frame}{Software for analysis}

\begin{itemize}
\item  \R : \emp{\url{https://www.R-project.org}}.
  \begin{itemize}
    \item \emp{RStudio}: \url{https://www.rstudio.com}.
  \end{itemize}    
\item \R\ \emp{packages} (\emp{\url{https://cran.R-project.org}}):
  \begin{itemize}
  \item \emp{skum} (Broström, 2016) data from \emp{DDB}.
  \item \emp{rmarkdown} (RStudio Team, 2016).
  \item \emp{knitr} (Xie, 2016).
  \item \emp{eha} (Broström, 2016).
  \item \emp{xtable} (Dahl, 2016).
\end{itemize}            
\end{itemize}

@@

@@

\begin{itemize}
\item \emp{Improved} survival characterised all {\sc Hisclass}es.
\item There are \emp{gender differences} in the social pattern
\item No difference in the social pattern in the two studied age groups.
\item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
\end{itemize}

>>>>

\begin{frame}{Questions}

@@

\begin{frame}{Where?}

\includegraphics[height=5cm]{myfigs/geo_we.pdf}\includegraphics[height=7cm]{myfigs/sweden.pdf}
\end{frame}

\begin{frame}{Here!}

\begin{columns}

\begin{column}{0.5\textwidth}

\includegraphics[height=7cm]{myfigs/skum_latest.pdf}
\end{column}

\begin{column}{0.5\textwidth}

Skellefteå:

\item Small town, heavy industry in the nineteenth century.

\end{itemize}

Umeå:

\item Not so small town, older. Center for administration, education, and military.

\end{itemize}

Rural areas:

\item Very similar: Small farms dominating.

\end{itemize}

\end{column}
\end{columns}
%\includegraphics[height=7cm]{myfigs/sweden.pdf}

@@

\begin{itemize}
\item \emp{Swedish church book} records (1851--1950).
  \begin{itemize}
\item digitized by the \emp{Demographic Data Base} at Umeå University,
\begin{itemize}
  % \item \emp{\sc Popum} and
   \item \emp{\sc Poplink}
\end{itemize}
\item  \url{http://www.cedar.umu.se/}
\end{itemize}
\item The \emp{Linnaeus} data base (1960--2013)
  \begin{itemize}
\item Census data 1960, 1965, \ldots, 1990
\item Population registers ({\sc Lisa}, 1986--2013)
  \begin{itemize}
    \item \emp{Income}
    \item \emp{Education}
  \end{itemize}
 \item Death information from \emp{National Board of Health and Welfare}
   \begin{itemize}
\item Death dates from 1 January 1961 to 31 December 2013.
  \item \emp{Causes of death} ({\sc Ulorsak}, ``main cause of death'')
   \end{itemize}
\end{itemize}
\end{itemize}

@@

load("data/vb.rda")
vbd <- vb[vb$event == 1 & vb$dataset != "poplink", c("birthdate", "exit", "sex")]
x <- vbd$birthdate + vbd$exit
y <- vbd$exit
##par(mfrow = c(1, 2))
plot(x[vbd$sex == "male"], y[vbd$sex == "male"], type = "n",
     pch = ".", col = "blue", xlab = "Year", ylab = "Age", las = 2,
     axes = FALSE)
axis(1, at = c(seq(1960, 1990, by = 5), 2002, 2007, 2014), las = 2)
axis(2, at = c(40, 65, 90))
axis(4, at = c(40, 65, 90))
for (i in seq(1960, 1990, by = 5) + (365 - 61) / 365){
    abline(v = i, lwd = 2)
}
for (i in 2002:2013){
    abline(v = i, lwd = 2, col = "black")
}
abline(h = c(40, 65, 90), lty = 2)
points(x[vbd$sex == "female"], y[vbd$sex == "female"], type = "n",
     pch = ".", col = "red")
box()
@

\begin{columns}
\begin{column}{0.6\textwidth}
<<picklexus2,fig.width=6,fig.height = 5.9>>=

@@

@@

\begin{frame}{Answers(?)}

\begin{enumerate}
\item Has \emp{inequality in mortality} between \emp{social classes}
  increased?
\begin{itemize}
\item Not really
\end{itemize}
%%\item Is the mortality transition initiated by a certain social class?

\item Are there any \emp{gender differences} in the effect of social position?
\begin{itemize}
\item Yes
\end{itemize}

\item Is social position \emp{equally important} among the \emp{retired} population as
  for those in \emp{working age}?
\begin{itemize}
\item Not really
\end{itemize}

\item Is there a difference between classes in mortality from
  \emp{cardiovascular} diseases and \emp{cancers} respectively?
\begin{itemize}
\item Yes
\end{itemize}
\item Are there large differences in survival depending on level of
  \emp{education} and \emp{income} categories?
\begin{itemize}
\item Yes, definitely.
\end{itemize}

@@

\begin{itemize}
\item \emp{Improved} survival characterised all {\sc Hisclass}es.
\item There are \emp{gender differences} in the social pattern
\item No difference in the social pattern in the two studied age groups.
\item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
\item \emp{Education} and \emp{income} are the \emp{most important}
  variables during late
  time period.
\end{itemize}

\begin{frame}{Software for report writing and analysis}

\begin{itemize}
\item  \R : \emp{\url{https://www.R-project.org}}.
  \begin{itemize}
    \item \emp{RStudio}: \url{https://www.rstudio.com}.
  \end{itemize}
\item \R\ \emp{packages} (\emp{\url{https://cran.R-project.org}}):
  \begin{itemize}
  \item \emp{skum} (Broström, 2016) data from \emp{DDB}.
  \item \emp{rmarkdown} (RStudio Team, 2016).
  \item \emp{knitr} (Xie, 2016).
  \item \emp{eha} (Broström, 2016).
  \item \emp{xtable} (Dahl, 2016).
\end{itemize}

\item \emp{Version control} (reproducibility)
  \begin{itemize}
    \item \emp{git}: \url{http://git}
    \item \emp{gitHub}: \url{http://github.com/goranbrostrom/}

\end{itemize}
\end{itemize}
